Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-06-19 _tex/manuscript.tex
Built site for gh-pages

@@

When treating head and neck squamous cell carcinomas (HNSCC) with
radiotherapy or surgery, not only the primary tumor and the clinically
detected lymph node metastases are targeted. Per current guidelines,
large volumes of the neck - lymph node levels (LNLs) - are part of the
the elective clinical target volume (CTV-N)
\citep{gregoire_ctbased_2003, gregoire_delineation_2014, gregoire_delineation_2018, eisbruch_intensitymodulated_2002, biau_selection_2019, chao_determination_2002, vorwerk_guidelines_2011, ferlito_elective_2009}.
This is to minimize the risk of regional recurrences due to missed
microscopic disease which cannot be detected by current in-vivo imaging
modalities such as computed tomography (CT), magnetic resonance imaging
(MRI), or positron emission tomography (PET).

@@

We have collected a detailed dataset of 833 patients with newly
diagnosed oropharyngeal squamous cell carcinomas. It reports the
involvement of every patient individually and per lymph node level in
tabular form, in addition to other clinico-pathological information such
as age, T-category, and HPV p16 status.

@@

\caption{}\label{T_42bd2}\tabularnewline
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Centre Léon Bérard & 325 & 60 & 100\% & 19\% & 69\% & 18\% \\
Inselspital Bern & 74 & 61 & 100\% & 18\% & 66\% & 14\% \\
University Hospital Zurich & 287 & 66 & 26\% & 18\% & 52\% & 31\% \\
Vall d\textquotesingle Hebron Barcelona Hospital & 147 & 58 & 5\% & 21\%
& 34\% & 34\% \\

@@

The entire data, including additional patients with tumors in other
primary locations than the oropharynx, is publicly available: It may be
\href{https://lyprox.org/patients/dataset}{downloaded from LyProX} where
it can be interactively explored too,
\href{https://github.com/rmnldwg/lydata}{from GitHub},
\href{https://zenodo.org/search?q=lydata}{from zenodo}, or via the
\emph{Data-in-Brief} publications \citet{ludwig_dataset_2022} and
\citet{ludwig_multicentric_2023}. Although these publications do not
include the most recent dataset addition from Vall d'Hebron Barcelona
Hospital.

@@

It is possible to provide our model with multiple different diagnostic
modalities, each being characterized by different pairs of sensitivity
and specificity. However, we instead chose to combine them into a single
``consensus'' diagnosis before parameter inference. We opted for this
because the literature values of sensitivity and specificity
\citep{debondt_detection_2007} of imaging modalities like MRI and CT do
not plausibly match some of our observations: In the USZ cohort, 78\% of
OPSCC patients where diagnosed with ipsilateral LNL II involvement via
diagnostic imaging. This is virtually impossible with sensitivities
around 80\% and specificities lower than 100\%.

Our pre-training consensus was formed by considering all reported
diagnostic information for a particular patient and LNL. When conflicts
arose, we computed the \emph{most likely} true state of involvement
using the literature sensitivity and specificity values
\citep{debondt_detection_2007}.

@@

For parameter inference, we used the Python package
\href{https://emcee.readthedocs.io/en/stable/}{\texttt{emcee}}
\citep{foreman-mackey_emcee_2013}. It implements efficient MCMC sampling
algorithms that employ multiple parallel samplers for affine invariance
and better performance on multi-core CPUs. The
\href{https://emcee.readthedocs.io/en/stable/}{\texttt{emcee}} library
was provided with the likelihood implemented by our
\href{https://lymph-model.readthedocs.io/en/stable/}{\texttt{lymph-model}}
Python package.

@@

\caption{}\label{T_c1c15}\tabularnewline
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Mid. ext. probability & 8.13\% & ± 0.60\% \\
ipsi: T ➜ II & 35.45\% & ± 1.75\% \\
ipsi: T ➜ III & 5.59\% & ± 0.82\% \\
ipsi: T ➜ IV & 0.91\% & ± 0.21\% \\
contra: T ➜ II & 2.64\% & ± 0.35\% \\
contra: T ➜ III & 0.16\% & ± 0.09\% \\
contra: T ➜ IV & 0.21\% & ± 0.10\% \\
Mixing ⍺ & 21.01\% & ± 3.24\% \\
II ➜ III & 13.46\% & ± 1.95\% \\
III ➜ IV & 15.74\% & ± 2.24\% \\
late T-cat. binom. prob. & 45.35\% & ± 2.57\% \\

@@

@@

Figure~\ref{fig-model-risks} shows this predicted risk of occult disease
in contralateral LNL II for four interesting combinations of these three
risk factors. This risk is computed \emph{given} a CT diagnosis
assessing the per-LNL clinical involvement for which we assumed a
sensitivity of 81\% and a specificity of 76\%
\citep{debondt_detection_2007}.

@@

\begin{itemize}
\tightlist
\item
  conceptually extended model to cover both sides of neck
\item
  explicitly models midline extension (also allows for easy
  marginalization)
\item
  midline extension intuitively causes contralateral spread to become
  more similar to ipsilateral spread
\end{itemize}

@@

\caption{}\label{T_78884}\tabularnewline
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
early & 0 & False & 0 & 0.00 & 1 & 1.16 & 0 & 0.00 & 0 & 0.00 & 86 \\
early & 0 & True & 0 & 0.00 & 1 & 10.00 & 1 & 10.00 & 0 & 0.00 & 10 \\
early & 1 & False & 1 & 0.53 & 11 & 5.82 & 2 & 1.06 & 1 & 0.53 & 189 \\
early & 1 & True & 1 & 11.11 & 2 & 22.22 & 0 & 0.00 & 0 & 0.00 & 9 \\
early & ≥ 2 & False & 1 & 0.96 & 15 & 14.42 & 3 & 2.88 & 3 & 2.88 &
104 \\
early & ≥ 2 & True & 0 & 0.00 & 3 & 30.00 & 4 & 40.00 & 1 & 10.00 &
10 \\
advanced & 0 & False & 0 & 0.00 & 2 & 5.88 & 0 & 0.00 & 0 & 0.00 & 34 \\
advanced & 0 & True & 0 & 0.00 & 3 & 12.50 & 0 & 0.00 & 0 & 0.00 & 24 \\
advanced & 1 & False & 0 & 0.00 & 3 & 4.55 & 0 & 0.00 & 0 & 0.00 & 66 \\
advanced & 1 & True & 1 & 1.64 & 18 & 29.51 & 5 & 8.20 & 1 & 1.64 &
61 \\
advanced & ≥ 2 & False & 4 & 5.26 & 16 & 21.05 & 7 & 9.21 & 4 & 5.26 &
76 \\
advanced & ≥ 2 & True & 3 & 3.80 & 46 & 58.23 & 18 & 22.78 & 8 & 10.13 &
79 \\

>>>>

\abstract{In current practice, large parts of the contralateral lymph
system are included in the elective clinical target volume (CTV-N) of
oropharyngeal squamous cell carcinoma (OPSCC) patients. Guidelines on
how to define the CTV-N are mostly based on prevalences of nodal
involvement in the different lymph node levels (LNLs) and do not
consider the detailed, individual diagnosis of a patient.

In this work, we present a probabilistic model for bilateral lymphatic
tumor progression in OPSCC. Its aim is to predict the personal risk for
occult disease in any LNL, given the patient's clincal diagnosis,
T-category, and lateralization of the primary tumor. The model
represents each of the LNLs I, II, III, IV, V, and VII of both sides of
the neck as a hidden binary random variable that are connected to the
tumor and among each other via arcs that correspond to spread
probabilities per time-step. These spread probability rates are learned
via Markov chain Monte Carlo (MCMC) sampling from a dataset of 833 OPSCC
patients.

The model is able to describe the data well and may in the future be
used to only include LNLs in a CTV-N that show a risk above a specific
threshold for that particular patient's diagnosis. E.g., it predicts the
risk for occult disease in contralateral LNL II to be below 5\% unless
the tumor crosses the mid-sagittal line.}

@@

When treating head and neck squamous cell carcinomas (HNSCC) with
radiotherapy or surgery, not only the primary tumor and the clinically
detected lymph node metastases are targeted. Per current guidelines,
large volumes of the neck - lymph node levels (LNLs) - are part of the
the elective clinical target volume (CTV-N)
\citep{gregoire_ct-based_2003, gregoire_delineation_2014, gregoire_delineation_2018, eisbruch_intensity-modulated_2002, biau_selection_2019, chao_determination_2002, vorwerk_guidelines_2011, ferlito_elective_2009}.
This is to minimize the risk of regional recurrences due to missed
microscopic disease which cannot be detected by current in-vivo imaging
modalities such as computed tomography (CT), magnetic resonance imaging
(MRI), or positron emission tomography (PET).

@@

Originally, the only available data of detailed - i.e.~per patient and
per LNL - lymphatic involvement was a reconstructed table from the
findings in \citet{sanguineti_defining_2009}, which included only
ipsilateral nodal involvement for the LNLs I to IV.

To be able to create models for lymphatic tumor progression that include
more LNLs and the contralateral side, we have collected a detailed
dataset of 833 patients with newly diagnosed oropharyngeal squamous cell
carcinomas. It reports the involvement of every patient individually and
per lymph node level in tabular form, in addition to other
clinico-pathological information such as age, T-category, and HPV p16
status.

@@

\caption{}\label{T_331cf}\tabularnewline
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Centre Léon Bérard & 325 & 60 & 100\% & 19\% & 69\% & 18\% \\
Inselspital Bern & 74 & 61 & 100\% & 18\% & 66\% & 14\% \\
University Hospital Zurich & 287 & 66 & 26\% & 18\% & 52\% & 31\% \\
Vall d\textquotesingle Hebron Barcelona Hospital & 147 & 58 & 5\% & 21\%
& 34\% & 34\% \\

@@

The entire data, including additional patients with tumors in other
primary locations than the oropharynx, is publicly available: It may be
\href{https://lyprox.org/patients/dataset}{downloaded from LyProX} where
it can be interactively explored too,
\href{https://github.com/rmnldwg/lydata}{from GitHub},
\href{https://zenodo.org/search?q=lydata}{from zenodo}, or via the
\emph{Data-in-Brief} publications \citet{ludwig_dataset_2022} and
\citet{ludwig_multi-centric_2023}. Although these publications do not
include the most recent dataset addition from Vall d'Hebron Barcelona
Hospital.

@@

It is possible to provide our model with multiple different diagnostic
modalities, each being characterized by different pairs of sensitivity
and specificity. However, we instead chose to combine them into a single
``consensus'' diagnosis before parameter inference. We opted for this
because the literature values of sensitivity and specificity
\citep{de_bondt_detection_2007, kyzas_18f-fluorodeoxyglucose_2008} of
imaging modalities like MRI and CT do not plausibly match some of our
observations: In the USZ cohort, 78\% of OPSCC patients where diagnosed
with ipsilateral LNL II involvement via diagnostic imaging. This is
virtually impossible with sensitivities around 80\% and specificities
lower than 100\%.

Our pre-training consensus was formed by considering all reported
diagnostic information for a particular patient and LNL. When conflicts
arose, we computed the \emph{most likely} true state of involvement
using the literature sensitivity and specificity values
\citep{de_bondt_detection_2007, kyzas_18f-fluorodeoxyglucose_2008}.

@@

For parameter inference, we used the Python package
\href{https://emcee.readthedocs.io/en/stable/}{\texttt{emcee}}
\citep{foreman-mackey_emcee_2013}. It implements efficient MCMC sampling
algorithms that employ multiple parallel samplers for affine invariance
and better performance on multi-core CPUs. The sample proposal
algirothms used by us are based on differential evolution moves
\citep{ter_braak_differential_2008, nelson_run_2013}. The
\href{https://emcee.readthedocs.io/en/stable/}{\texttt{emcee}} library
was provided with the likelihood implemented by our
\href{https://lymph-model.readthedocs.io/en/stable/}{\texttt{lymph-model}}
Python package.

@@

\caption{}\label{T_d4a82}\tabularnewline
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Mid. ext. probability & 8.13\% & ± 0.60\% \\
ipsi: T ➜ II & 35.45\% & ± 1.75\% \\
ipsi: T ➜ III & 5.59\% & ± 0.82\% \\
ipsi: T ➜ IV & 0.91\% & ± 0.21\% \\
contra: T ➜ II & 2.64\% & ± 0.35\% \\
contra: T ➜ III & 0.16\% & ± 0.09\% \\
contra: T ➜ IV & 0.21\% & ± 0.10\% \\
Mixing ⍺ & 21.01\% & ± 3.24\% \\
II ➜ III & 13.46\% & ± 1.95\% \\
III ➜ IV & 15.74\% & ± 2.24\% \\
late T-cat. binom. prob. & 45.35\% & ± 2.57\% \\

@@

Also, the discrepancy in midline extension prevalence between early and
advanced T-category is particularly pronounced in oropharyngeal SCC
patients. For example, in oral cavity SCC, the midline extension only
increases from 15.4\% (20 out of 130) to 33.3\% (13 out of 39).

@@

Figure~\ref{fig-model-risks} shows this predicted risk of occult disease
in contralateral LNL II for four interesting combinations of these three
risk factors. This risk is computed \emph{given} a CT diagnosis
assessing the per-LNL clinical involvement for which we assumed a
sensitivity of 81\% and a specificity of 76\%
\citep{de_bondt_detection_2007}.

@@

\subsection{Summary}\label{summary}

In this work we present a formalism to model the ipsi- and contralateral
lymphatic involvement of oropharyngeal SCC patients. An ipsilateral
model has been previously developed and published
\citep{ludwig_hidden_2021, ludwig_modelling_2023}. Based on this, we
introduce an extension that leaves the ipsilateral model untouched, but
extends it to the contralateral side in an intuitive and comprehensible
manner.

The model's performance w.r.t. its ability to describe the data on
contralateral nodal involvement was evaluated based on a dataset of 833
patients from four institutions. To the best of our knowledge, this is
the first time lymphatic tumor progression has been modelled in such
detail and to such an extent. While there was work on regional tumor
spread, these models were conceptually different, more limited in the
LNLs for which they make predictions, and not trained with real patient
data \citep{benson_markov_2006, jung_development_2017}.

The model takes the clincal diagnosis of the lymph node levels, the
primary tumor's T-category, as well as lateralization into account when
predicting the personalized risk for occult disease in any LNL of
interest. Owing to the relatively few parameters, the model is highly
interpretable and every parameter can be intuitively explained, while
still offering good accuracy in its predictions.

\subsection{Implications for Contralateral Elective Nodal
Treatment}\label{implications-for-contralateral-elective-nodal-treatment}

The predictions of this model are being used to inform a clinical trial
on volume deescalation at the University Hospital Zurich {[}citation
needed{]}. When accepting a 5\% risk of occult disease in any given LNL,
the model suggests the following contralateral elective irradiation
(assuming the respective LNL appears clinically healthy):

\begin{itemize}
\tightlist
\item
  LNL II when the tumor extends over the mid-sagittal line
\item
  LNL III when LNL II is clinically, the tumor is of advanced
  T-category, crosses the midline, and the ipsilateral neck shows
  involvement of at least LNL II
\item
  LNL IV only for severe ipsilateral involvement, advanced T-category
  with obvious midline extension and contralateral upstream involvement
\end{itemize}

@@

\caption{}\label{T_f8f73}\tabularnewline
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
early & 0 & False & 0 & 0.00 & 1 & 1.16 & 0 & 0.00 & 0 & 0.00 & 86 \\
early & 0 & True & 0 & 0.00 & 1 & 10.00 & 1 & 10.00 & 0 & 0.00 & 10 \\
early & 1 & False & 1 & 0.53 & 11 & 5.82 & 2 & 1.06 & 1 & 0.53 & 189 \\
early & 1 & True & 1 & 11.11 & 2 & 22.22 & 0 & 0.00 & 0 & 0.00 & 9 \\
early & ≥ 2 & False & 1 & 0.96 & 15 & 14.42 & 3 & 2.88 & 3 & 2.88 &
104 \\
early & ≥ 2 & True & 0 & 0.00 & 3 & 30.00 & 4 & 40.00 & 1 & 10.00 &
10 \\
advanced & 0 & False & 0 & 0.00 & 2 & 5.88 & 0 & 0.00 & 0 & 0.00 & 34 \\
advanced & 0 & True & 0 & 0.00 & 3 & 12.50 & 0 & 0.00 & 0 & 0.00 & 24 \\
advanced & 1 & False & 0 & 0.00 & 3 & 4.55 & 0 & 0.00 & 0 & 0.00 & 66 \\
advanced & 1 & True & 1 & 1.64 & 18 & 29.51 & 5 & 8.20 & 1 & 1.64 &
61 \\
advanced & ≥ 2 & False & 4 & 5.26 & 16 & 21.05 & 7 & 9.21 & 4 & 5.26 &
76 \\
advanced & ≥ 2 & True & 3 & 3.80 & 46 & 58.23 & 18 & 22.78 & 8 & 10.13 &
79 \\
